---
title: "Run PNW Bat Analysis"
author: "DAU, modified from Trent VanHawkins"
date: "`r Sys.Date()`"
format: 
  html:
    lightbox: true
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(sf)
```

# PNW Bat Trend Analysis

## Step 0. Extract tables

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`00_ExportTables_alt.R`](src/00_ExportTables_alt.R) | Extracts database tables from MS Access (`.accdb`) to CSV using `RODBC`. | `data/raw/database/*.accdb` | `data/raw/tables/*.csv` |

Databases are copied from the BatHub Box folders to `PNW_BatTrend/data/raw/databases`. The first script, `00_ExportTables_alt.R`, will export required tables from the databases below into the `data/raw/tables` folder.

### Databases Used in Analysis

Adjust source script, `00_ExportTables_alt.R`, if new data exists or filenames have changed.

| Database File | Purpose / Coverage | Date Last Modified |
|-------------------------|-------------------------------|----------------|
| `PNW_BatHub_Database_20260414.accdb` | Main BatHub database containing survey, site, and metadata records. | 8/26/2026 |
| `PNW_BatHub_Database_AcousticOutput_2016-2023.accdb` | Acoustic processing outputs for monitoring data collected from 2016-2023. | 2/4/2026 |
| `PNW_BatHub_Database_AcousticOutput_2024.accdb` | Acoustic processing outputs for 2024 monitoring data. | 4/13/2026 |
| `PNW_BatHub_Database_AcousticOutput_2025.accdb` | Acoustic processing outputs for 2025 monitoring data. | 4/13/2026 |
| `PNW_BatHub_Database_AcousticOutput_Idaho_2025.accdb` | Acoustic processing outputs for Idaho 2025 monitoring data. | 8/26/2026 |

```{r}
#| label: run00
#| eval: false

# already ran and exported tables
source(here("src/00_ExportTables_alt.R"))
```

## Step 1. Clean deployments

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`01_DeploymentsCleaning.R`](src/01_DeploymentsCleaning.R) | Cleans acoustic detector deployment records, extracts metadata, and exports surveyed locations. | `data/raw/tables/` (`tblDeployment.csv`, `tblPointLocation.csv`, `tblSite.csv`, `tluClutterType.csv`, `tluWaterBodyType.csv`) | `data/processed/detections/deployments_to2025.rds`, `data/raw/covariates/daymet/daymet_sites.csv` |

```{r}
#| label: run01
#| warning: false

source(here("src/01_DeploymentsCleaning.R"))
```

### Missing sites

```{r}
#| label: run01Missing
#| echo: false
#| message: false

missing_sites <- read_csv(here("data/processed/detections/sites_missing_covars.csv"), 
                               show_col_types = FALSE)
```

Missing data were saved at `data/processed/detections/sites_missing_covars.csv`.

```{r}
#| label: fig-missing-covars
#| echo: false
#| fig-cap: "Missing clutter or waterbody"

missing_sites %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  ggplot() +
  geom_sf(data = states_map, fill = "gray90", color = "black") +
  geom_sf(aes(color = factor(year)), alpha = 0.5) + 
  labs(color = "Year") +
  theme_bw() + 
  theme(panel.grid = element_blank())

```

### Available data

```{r}
#| label: fig-deployment-map
#| echo: false
#| fig-cap: "Deployment data across Idaho, Oregon, and Washington"

deployment %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  ggplot() +
  geom_sf(data = states_map, fill = "gray90", color = "black") +
  geom_sf(aes(color = state), alpha = 0.5) +
  facet_wrap(~year) +
  labs(color = "State") +
  theme_bw() +
  theme(legend.position = "bottom", 
        panel.grid = element_blank(),
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks = element_blank()) 
```

## Step 2. Compile spatial covariates

Requires layers to be pulled and stored in the `data/raw/` folder. See below for required files.

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`02_CompileSpatialCovariates.R`](src/02_CompileSpatialCovariates.R) | Extracts and aggregates spatial covariates (forest cover, karst, cliff/canyon, elevation, precipitation) for the NABat 10km grid. | `data/raw/covariates/`, `data/raw/batgrid/` | `data/processed/occurrence/batgrid_covars.shp` |

| Variable | Local File Path | Download Date | Source |
|-------------|------------------------------|-------------|----------------|
| nabat_covars | covariates/NABat_grid_covariates/ | 2026-08-31 | [2025 North American Bat Monitoring Program Knitted Grid and Multi-scale Grid Covariates](https://www.sciencebase.gov/catalog/item/67eafce2d34ed02007f83722) |
| conus10k | batgrid/conus_mastersample_10km_attributed.shp | 2026-08-31 | [2018 Grid-Based Sampling Frame](https://www.sciencebase.gov/catalog/item/5b7b563ae4b0f5d57884615b) |
| landfire_or | covariates/LandFire/LF2022_OR/LC22_EVT_230.tif | 2026-08-19 | [landfire](https://www.landfire.gov/vegetation/evt) |
| landfire_wa | covariates/LandFire/LF2022_WA/LC22_EVT_230.tif | 2026-08-19 | [landfire](https://www.landfire.gov/vegetation/evt) |
| landfire_id | covariates/LandFire/LF2022_ID/LC22_EVT_230.tif | 2026-08-19 | [landfire](https://www.landfire.gov/vegetation/evt) |

```{r}
#| label: run02
#| warning: false
#| eval: false
#| message: false

# already run -- need to check file names
source(here("src/02_CompileSpatialCovariates.R"))
```

Successful run of this script will output `batgrid_covars`, saved at `data/processed/occurrence/batgrid_covars.shp`.

## Step 3. Clean detections

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`03_DetectionsCleaning.R`](src/03_DetectionsCleaning.R) | Cleans and harmonizes species call identifications across all acoustic output CSVs (2016–2025). | `data/processed/detections/deployments_to2025.rds`, `data/raw/tables/calls_*.csv` | `data/processed/detections/detection_histories.rds` |

```{r}
#| label: run03
#| warning: false

source(here("src/03_DetectionsCleaning.R"))
```

## Step 4. Get Daymet

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`04_get_daymet.py`](file:///src/04_get_daymet.py) | Downloads daily Daymet weather data (tmin, dayl, prcp) for surveyed sites via `pydaymet`. | `data/raw/covariates/daymet/daymet_sites.csv` | `data/raw/covariates/daymet/daymet_output.csv` |

::: {.callout-caution icon="false"}
Run once, from a terminal (not the R console):

Requires `pandas` and `pydaymet` installed for whatever `python` resolves to on your system.

```         
python -m pip install pandas
```

```         
python -m pip install pydaymet
```

```         
python src/04_get_daymet.py
```
:::

## Step 5. Prepare data for modeling

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`05_detections_modprep.R`](src/05_detections_modprep.R) | Merges cleaned species detections with climate covariates and formats nightly survey replicates. | `data/processed/detections/detection_histories.rds`, `data/raw/covariates/daymet/daymet_output.csv` | `data/processed/detections/nw_nights.rds` |

```{r}
#| label: run05
#| warning: false
#| message: false

source(here("src/05_detections_modprep.R"))
```

```{r}
#| label: fig-spatial-reps
#| fig-cap: "Visualize how many spatial replicates in each cell"

ggplot(spatial_reps_summary, aes(x = spatial_reps)) + 
    geom_bar(color = "black", 
             fill = "darkgray") + 
    facet_wrap(~year) +
    theme_bw() + theme(panel.grid = element_blank())
```

## Step 6. Prepare covariate data for occupancy model

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`06_occurrence_modprep.R`](src/06_occurrence_modprep.R) | Finalizes grid-level occurrence covariate tables and joins historical survey status. | `data/processed/occurrence/batgrid_covars.shp`, `data/processed/detections/nw_nights.rds` | `data/processed/occurrence/nw_grid.rds` |

```{r}
#| label: run06
#| warning: false
#| eval: false

# don't have data yet
source(here("src/06_occurrence_modprep.R"))
```

## Step 7. Dynamic occupancy modeling

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`07_dynocc_autolog.R`](src/07_dynocc_autolog.R) | Formats Stan data matrices and fits multi-species autologistic dynamic occupancy models across priors. | `data/processed/occurrence/nw_grid.rds`, `data/processed/detections/nw_nights.rds` | `data/processed/results/stan/stan_data_*.rds`, `design_matrices.rds`, `index_keys.rds`, fits in `data/processed/results/stan/fits/` |
| [`07_dynoccMod.R`](src/07_dynoccMod.R) | Formats Stan data matrices and fits multi-species autologistic dynamic occupancy models across priors. | `data/processed/occurrence/nw_grid.rds`, `data/processed/detections/nw_nights.rds` | `data/processed/results/stan/stan_data_*.rds`, `design_matrices.rds`, `index_keys.rds`, fits in `data/processed/results/stan/full/fits/` |
| [`dynocc_forward.stan`](src/dynocc_autologistic.stan) |  | N/A (Compiled by `cmdstanr`) | N/A |
| [`dynocc_autologistic.stan`](src/dynocc_autologistic.stan) | Stan model file implementing a first-order autologistic dynamic occupancy model with time-varying survival ($\phi_t$), colonization ($\gamma_t$), and logit-linear detection and initial occurrence covariates. | N/A (Compiled by `cmdstanr`) | N/A |

### Step 7a. Prepare data for dynamic occupancy model in Stan

Prepare detection and covariate data for dynamic occupancy model in Stan (**`dynocc_autologistic.stan`**). Outputs a species-level list of data objects and fits the Stan model for each species. The script produces the following outputs:

``` text
data/processed/results/stan/
├── stan_data_weakly_informative.rds
├── design_matrices.rds
├── index_keys.rds
└── fits/
    ├── laci_stanfit_weakly_informative.rds
    ├── lano_stanfit_weakly_informative.rds
    ├── myev_stanfit_weakly_informative.rds
    ├── ...
    └── coto_stanfit_weakly_informative.rds
```

```{r}
#| label: run07a
#| warning: false
#| eval: false

# don't have data yet
source(here("src/07_dynocc_autolog.R"))
```

### Step 7b. Prepare data for dynamic occupancy model in Stan

**Potentially an older version?** Prepare detection and covariate data for dynamic occupancy model in Stan (**`TV_dynocc_forward.stan`**). Outputs a species-level list of data objects and fits the Stan model for each species. The script produces the following outputs:

``` text
data/processed/results/stan/
├── stan_data_weakly_informative.rds
├── stan_data_weakly_informative_spatial.rds
├── design_matrices.rds
├── index_keys.rds
└── full/
  └── fits/
    ├── laci_stanfit_weakly_informative.rds
    ├── lano_stanfit_weakly_informative.rds
    ├── myev_stanfit_weakly_informative.rds
    ├── ...
    └── coto_stanfit_weakly_informative.rds
```

::: {.callout-caution icon="false"}
The `stan_data_weakly_informative.rds` overwrites the same file created in Step 7a. The files contain slightly different information.
:::

```{r}
#| label: run07b
#| warning: false
#| eval: false

# don't have data yet
source(here("src/07_dynoccMod.R"))
```

## Step 8. Check sensitivity of model?

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`08_dynoccSensitivity.R`](src/08_dynoccSensitivity.R) | Runs sensitivity analysis by fitting models excluding Idaho survey data. | `data/processed/occurrence/nw_grid.rds`, `data/processed/detections/nw_nights.rds` | Sensitivity design matrices and fits in `data/processed/results/stan/fits/` |

Prepare detection and covariate data for dynamic occupancy model in Stan (`dynocc_autologistic.stan`). Outputs a species-level list of data objects and fits the Stan model for each species. Importantly, this performs a sensitivity analysis by excluding Idaho and saved all output with a `_sensitivity` suffix. The script produces the following outputs:

``` text
data/processed/results/stan/
├── stan_data_weakly_informative_sensitivity.rds
├── design_matrices_sensitivity.rds
├── index_keys_sensitivity.rds
└── fits/
    ├── laci_stanfit_weakly_informative_sensitivity.rds
    ├── lano_stanfit_weakly_informative_sensitivity.rds
    ├── myev_stanfit_weakly_informative_sensitivity.rds
    ├── ...
    └── coto_stanfit_weakly_informative_sensitivity.rds
```

```{r}
#| label: run08
#| warning: false
#| eval: false

# don't have data yet
source(here("src/08_dynoccSensitivity.R"))
```

## Step 9. stan Diagnotics

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`09_stanDiagnostics.R`](src/09_stanDiagnostics.R) | Evaluates MCMC convergence, computes posterior predictive checks, calculates Moran's I correlograms (10–50 km), estimates regional occupancy trajectories/slopes, and outputs multi-species and individual species figures. | Stan fits in `data/processed/results/stan/fits/` | Diagnostic tables (`all_res_stan.rds`, `trend_full_stan.rds`, etc.) and plots in `data/processed/results/stan/diagnostics/` |

```{r}
#| label: run09
#| warning: false
#| eval: false

# don't have data yet
source(here("src/09_stanDiagnostics.R"))
```

```{r}
#| label: run09a
#| warning: false
#| eval: false

# don't have data yet
source(here("src/09a_stanDiagnostics.R"))
```

## Step 10. stan Maps

| File | Function | Input Files | Output Files |
|:-----------------|:-----------------|:-----------------|:-----------------|
| [`10_stanMaps.R`](src/10_stanMaps.R) | Projects posterior parameter distributions across the Northwest grid to create regional occupancy prediction maps and uncertainty plots. | Stan fits in `data/processed/results/stan/fits/`, `nw_grid.rds` | Spatial prediction maps in `data/processed/results/stan/maps/` |

```{r}
#| label: run10
#| warning: false
#| eval: false

# don't have data yet
source(here("src/10_stanMaps.R"))
```
