---
title: "Run PNW Bat Analysis"
author: "DAU, modified from Trent VanHawkins"
date: "`r Sys.Date()`"
format: 
  html:
    lightbox: true
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(sf)
```

# PNW Bat Trend Analysis

## Extract tables

Databases are copied from the BatHub Box folders to `PNW_BatTrend/data/raw/databases`. The first script, `00_ExportTables_alt.R`, will export required tables from the databases below into the `data/raw/tables` folder. 

### Databases Used in Analysis

Adjust source script, `00_ExportTables_alt.R`, if new data exists or filenames have changed. 

| Database File | Purpose / Coverage | Date Last Modified |
|---------------|-------------------|------|
| `PNW_BatHub_Database_20260414.accdb` | Main BatHub database containing survey, site, and metadata records. | 8/26/2026 |
| `PNW_BatHub_Database_AcousticOutput_2016-2023.accdb` | Acoustic processing outputs for monitoring data collected from 2016-2023. |2/4/2026 |
| `PNW_BatHub_Database_AcousticOutput_2024.accdb` | Acoustic processing outputs for 2024 monitoring data. | 4/13/2026 |
| `PNW_BatHub_Database_AcousticOutput_2025.accdb` | Acoustic processing outputs for 2025 monitoring data. | 4/13/2026 |
| `PNW_BatHub_Database_AcousticOutput_Idaho_2025.accdb` | Acoustic processing outputs for Idaho 2025 monitoring data. | 8/26/2026 |

```{r}
#| label: run00
#| eval: false
source(here("src/00_ExportTables_alt.R"))
```

## Clean deployments

```{r}
#| label: run01
#| warning: false

source(here("src/01_DeploymentsCleaning.R"))
```

### Missing sites

```{r}
#| label: run01Missing
#| echo: false
#| message: false

missing_sites <- read_csv(here("data/processed/detections/sites_missing_covars.csv"), 
                               show_col_types = FALSE)
```

Missing data were saved at `data/processed/detections/sites_missing_covars.csv`. 

```{r}
#| label: fig-missing-covars
#| echo: false
#| fig-cap: "Missing clutter or waterbody"

missing_sites %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  ggplot() +
  geom_sf(data = states_map, fill = "gray90", color = "black") +
  geom_sf(aes(color = factor(year)))

```

### Available data

```{r}
#| label: fig-deployment-map
#| echo: false
#| fig-cap: "Deployment data across Idaho, Oregon, and Washington"

deployment %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  ggplot() +
  geom_sf(data = states_map, fill = "gray90", color = "black") +
  geom_sf(aes(color = state)) +
  facet_wrap(~year) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(color = "State")
```

## Compile spatial covariates

Requires layers to be pulled and stored in the `data/raw/` folder. See below for required files. 

| Variable     | Local File Path                                                | Download Date | Source                                   |
|--------------|-------------------------------------------------------------------------|---------------|------------------------------------------|
| nabat_covars | covariates/NABat_grid_covariates/                              | *still needed*| *source link*                            |
| conus10k     | batgrid/complete_conus_mastersample_10km_attributed.shp        | *still needed*| *source link*                            |
| landfire_or  | covariates/LandFire/LF2022_OR/LC22_EVT_230.tif                 | 2026-08-19    | [landfire](https://www.landfire.gov/vegetation/evt)  |
| landfire_wa  | covariates/LandFire/LF2022_WA/LC22_EVT_230.tif                 | 2026-08-19    | [landfire](https://www.landfire.gov/vegetation/evt)  |
| landfire_id  | covariates/LandFire/LF2022_ID/LC22_EVT_230.tif                 | 2026-08-19    |[landfire](https://www.landfire.gov/vegetation/evt) |

```{r}
#| label: run02
#| warning: false
#| eval: false

# don't currently have all files
source(here("src/02_CompileSpatialCovariates.R"))
```

Successful run of this script will output `batgrid_covars`, saved at `data/processed/occurrence/batgrid_covars.shp`

## Clean detections

```{r}
#| label: run03
#| warning: false

source(here("src/03_DetectionsCleaning.R"))
```

# Get Daymet

**Fetches Daymet climate data → writes to `data/daymet/`**

Run once, from a terminal (not the R console):

Requires `pandas` and `pydaymet` installed for whatever `python` resolves to on your system.

```
python -m pip install pandas
```

```
python -m pip install pydaymet
```

```
python src/04_get_daymet.py
```

## Prepare data for modeling

Data are saved to `data/processed/nw_nights.rds`
```{r}
#| label: run05
#| warning: false
#| message: false

source(here("src/05_detections_modprep.R"))
```

```{r}
#| label: fig-spatial-reps
#| fig-cap: "Visualize how many spatial replicates in each cell"

ggplot(spatial_reps_summary, aes(x = spatial_reps)) + 
    geom_bar(color = "black", 
             fill = "darkgray") + 
    facet_wrap(~year) +
    theme_bw() + theme(panel.grid = element_blank())
```

## Prepare covariate data for occupancy model

```{r}
#| label: run06
#| warning: false
#| eval: false

# don't have data yet
source(here("src/06_occurrence_modprep.R"))
```

## Prepare data for dynamic occupancy model in Stan

Are we using both autolog? and Mod?

```{r}
#| label: run07a
#| warning: false
#| eval: false

# don't have data yet
source(here("src/07_dynocc_autolog.R"))
```

## Prepare data for dynamic occupancy model in Stan

Are we using both autolog? and Mod?

```{r}
#| label: run07b
#| warning: false
#| eval: false

# don't have data yet
source(here("src/07_dynoccMod.R"))
```

## Check sensitivity of model?

```{r}
#| label: run08
#| warning: false
#| eval: false

# don't have data yet
source(here("src/08_dynoccSensitivity.R"))
```

## stan Diagnotics

```{r}
#| label: run09
#| warning: false
#| eval: false

# don't have data yet
source(here("src/09_stanDiagnostics.R"))
```

## stan Maps

```{r}
#| label: run10
#| warning: false
#| eval: false

# don't have data yet
source(here("src/10_stanMaps.R"))
```