---
title: "Run PNW Bat Analysis"
author: "DAU, modified from Trent VanHawkins"
date: "`r Sys.Date()`"
format: 
  html:
    lightbox: true
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(sf)
```

# PNW Bat Trend Analysis

## Extract tables

Databases are copied from the BatHub Box folders to `PNW_BatTrend/data/raw/databases`. The first script, `00_ExportTables_alt.R`, will export required tables from the databases below into the `data/raw/tables` folder. 

### Databases Used in Analysis

Adjust source script, `00_ExportTables_alt.R`, if new data exists or filenames have changed. 

| Database File | Purpose / Coverage | Date Last Modified |
|---------------|-------------------|------|
| `PNW_BatHub_Database_20260414.accdb` | Main BatHub database containing survey, site, and metadata records. | 8/26/2026 |
| `PNW_BatHub_Database_AcousticOutput_2016-2023.accdb` | Acoustic processing outputs for monitoring data collected from 2016-2023. |2/4/2026 |
| `PNW_BatHub_Database_AcousticOutput_2024.accdb` | Acoustic processing outputs for 2024 monitoring data. | 4/13/2026 |
| `PNW_BatHub_Database_AcousticOutput_2025.accdb` | Acoustic processing outputs for 2025 monitoring data. | 4/13/2026 |
| `PNW_BatHub_Database_AcousticOutput_Idaho_2025.accdb` | Acoustic processing outputs for Idaho 2025 monitoring data. | 8/26/2026 |

```{r}
#| label: run00
#| eval: false
source(here("src/00_ExportTables_alt.R"))
```

## Clean deployments

```{r}
#| label: run01
#| warning: false

source(here("src/01_DeploymentsCleaning.R"))
```

### Missing sites

```{r}
#| label: run01Missing
#| echo: false
#| message: false

missing_sites <- read_csv(here("data/processed/detections/sites_missing_covars.csv"), 
                               show_col_types = FALSE)
```

Missing data were saved at `data/processed/detections/sites_missing_covars.csv`. 

```{r}
#| label: fig-missing-covars
#| echo: false
#| fig-cap: "Missing clutter or waterbody"

missing_sites %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  ggplot() +
  geom_sf(data = states_map, fill = "gray90", color = "black") +
  geom_sf(aes(color = factor(year)))

```

### Available data

```{r}
#| label: fig-deployment-map
#| echo: false
#| fig-cap: "Deployment data across Idaho, Oregon, and Washington"

deployment %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  ggplot() +
  geom_sf(data = states_map, fill = "gray90", color = "black") +
  geom_sf(aes(color = state)) +
  facet_wrap(~year) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(color = "State")
```